---
title: "R vs. Python: Data Science Comparison"
author: "By Tanner Bessette"
output:
  beamer_presentation:
    theme: "Frankfurt"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(hrbrthemes)
Athletes_df <- read_csv("/Users/tannerbessette/Desktop/SYE/athletes.csv")
Medals_df <- read_csv("/Users/tannerbessette/Desktop/SYE/medals.csv")
```


```{r}
Male_Athletes_df <- Athletes_df %>%
  filter(gender == "Male")

Female_Athletes_df <- Athletes_df %>%
  filter(gender == "Female")
```


```{r}
Male_Total_Medal_Counts <- Medals_df %>%
  filter(gender == "M") %>%
  group_by(country_code) %>%
  summarise(medal_totals = n())

Female_Total_Medal_Counts <- Medals_df %>%
  filter(gender == "W") %>%
  group_by(country_code) %>%
  summarise(medal_totals = n()) 
```


```{r, include=FALSE}
Male_Athletes_Count <- Male_Athletes_df %>%
  group_by(country_code) %>%
  summarise(athlete_count = n()) %>%
  arrange(desc(athlete_count))

# join the medals counts with the athletes count
Male_Athletes_Count <- left_join(
  Male_Athletes_Count, Male_Total_Medal_Counts, by = "country_code")

# Calculate max country counts and visualize the top 10 countries:
Male_Athletes_Count <- Male_Athletes_Count %>%
  filter(athlete_count > 128)

# Create the plot:
Male_Athletes_Count_barplot <- ggplot(data = Male_Athletes_Count) +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = athlete_count), fill = "lightblue", stat = "identity") +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = medal_totals), fill = "darkblue", stat = "identity") +
          labs(title = "Male Athletes and Medals by Country",
               x = "Country",
               y = "Male Athletes and Medal Counts") +
          theme_minimal() +
          theme(legend.position = "none")

print(Male_Athletes_Count_barplot)
```


```{r, include=FALSE}
Female_Athletes_Count <- Female_Athletes_df %>%
  group_by(country_code) %>%
  summarise(athlete_count = n()) %>%
  arrange(desc(athlete_count))

# join the medals counts with the athletes count
Female_Athletes_Count <- left_join(
  Female_Athletes_Count, Female_Total_Medal_Counts, by = "country_code")

# Calculate max country counts and visualize the top 10 countries:
Female_Athletes_Count <- Female_Athletes_Count %>%
  filter(athlete_count > 177)

# Create the plot:
Female_Athletes_Count_barplot <- ggplot(data = Female_Athletes_Count) +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = athlete_count), fill = "lightblue", stat = "identity") +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = medal_totals), fill = "darkblue", stat = "identity") +
          labs(title = "Female Athletes and Medals by Country",
               x = "Country",
               y = "Female Athletes and Medal Counts") +
          theme_minimal() +
          scale_fill_viridis_c() +
          theme(legend.position = "none")

print(Female_Athletes_Count_barplot)
```


```{r}
# Create new Sport_Category variable for heatmap that reduces 
# the amount of categories:
Medal_Counts <- Medals_df %>%
  mutate(Sport_Category = as.factor(case_when(
    discipline %in% c("3x3 Basketball", "Basketball") ~ "Basketball",
    discipline %in% c("Badminton", "Table Tennis", "Tennis") ~ "Racket Sports",
    discipline %in% c("Boxing", "Taekwondo", "Wrestling", "Judo") ~ "Combat Sports",
    discipline %in% c("Swimming", "Marathon Swimming", "Diving", "Artistic Swimming") ~
      "Swimming",
    discipline %in% c("Volleyball", "Beach Volleyball") ~ "Volleyball",
    discipline %in% c("Cycling BMX Freestyle", "Cycling BMX Racing", 
                      "Cycling Mountain Bike",
                 "Cycling Road", "Cycling Track") ~ "Cycling",
    discipline %in% c("Water Polo", "Surfing", "Sailing") ~ "Water Sports",
    discipline %in% c("Artistic Gymnastics", "Rhythmic Gymnastics", 
                      "Trampoline Gymnastics") ~ "Gymnastics",
    discipline == "Athletics" ~ "Athletics",
    discipline %in% c("Canoe Slalom", "Canoe Sprint", "Rowing") ~ "Canoe/Rowing",
    discipline == "Equestrian" ~ "Equestrian",
    discipline %in% c("Football", "Handball", "Hockey", "Rugby Sevens") ~ 
      "Team Sports",
    discipline %in% c("Archery", "Shooting", "Fencing") ~ "Precision Sports",
    discipline %in% c("Triathlon", "Modern Pentathlon") ~ "Athlons",
    discipline == "Weightlifting" ~ "Weightlifting",
    discipline == "Skateboarding" ~ "Skateboarding",
    discipline == "Breaking" ~ "Breakdancing",
    discipline == "Sport Climbing" ~ "Climbing",
    discipline == "Golf" ~ "Golf",
    TRUE ~ "Other"  # default if any sport doesn't match
  )))
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
# group by our new Sport_Category variable before counting number of medals
Medal_Counts <- Medal_Counts  %>%
  group_by(country_code, Sport_Category) %>%
  mutate(count = n()) %>%
  arrange(desc(count))

# Only keep countries who got at least 10 total medals:
# (because heatmap was getting sloppy)
Medal_Counts <- Medal_Counts %>%
  group_by(country_code) %>%
  mutate(total_count = n()) %>%
  filter(total_count >= 10)

# Heatmap 
ggplot(Medal_Counts, aes(x = Sport_Category, 
                         y = fct_reorder(country_code, total_count),
                         fill= count)) + 
  geom_tile() +
  labs(title = "Medal Count by Country and Sport",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count") +
  coord_fixed()
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
# Only keep countries who got at least 10 total medals:
# (because heatmap was getting sloppy)
# group by our new Sport_Category variable before counting number of medals
Male_Medal_Counts <- Medal_Counts  %>%
  filter(gender == "M") %>%
  group_by(country_code, Sport_Category) %>%
  mutate(count = n()) %>%
  arrange(desc(count))

Male_Medal_Counts <- Male_Medal_Counts %>%
  group_by(country_code) %>%
  mutate(total_count = n()) %>%
  filter(total_count >= 10)

# Heatmap 
ggplot(Male_Medal_Counts, aes(x = Sport_Category, 
                              y = reorder(country_code, total_count), 
                              fill= count)) + 
  geom_tile() +
  labs(title = "Medal Count by Country and Sport (Male)",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count") +
  coord_fixed()
```


```{r, warning=FALSE, message=FALSE, include=FALSE}
# Only keep countries who got at least 10 total medals:
# (because heatmap was getting sloppy)
# group by our new Sport_Category variable before counting number of medals
Female_Medal_Counts <- Medal_Counts  %>%
  filter(gender == "W") %>%
  group_by(country_code, Sport_Category) %>%
  mutate(count = n()) %>%
  arrange(desc(count))

Female_Medal_Counts <- Female_Medal_Counts %>%
  group_by(country_code) %>%
  mutate(total_count = n()) %>%
  filter(total_count >= 10)

# Heatmap 
ggplot(Female_Medal_Counts, aes(x = Sport_Category, 
                              y = reorder(country_code, total_count), 
                              fill= count)) + 
  geom_tile() +
  labs(title = "Medal Count by Country and Sport (Female)",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count") +
  coord_fixed()
```



```{r, include=FALSE}
Male_Medal_Prop <- Male_Medal_Counts %>%
  group_by(Sport_Category) %>%
  mutate(total_sport_medals = sum(count)) %>%
  mutate(prop_sport_medals = count / total_sport_medals)

# Heatmap 
ggplot(Male_Medal_Prop, aes(x = Sport_Category, 
                              y = reorder(country_code, total_count), 
                              fill= prop_sport_medals)) + 
  geom_tile() +
  labs(title = "Proportion of Medals Won for Each Sport (Male)",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Proportion") +
  coord_fixed()
```

```{r, include=FALSE}
Female_Prop <- Female_Medal_Counts %>%
  group_by(Sport_Category) %>%
  mutate(total_sport_medals = sum(count)) %>%
  mutate(prop_sport_medals = count / total_sport_medals)

# Heatmap 
ggplot(Female_Prop, aes(x = Sport_Category, 
                              y = reorder(country_code, total_count), 
                              fill= prop_sport_medals)) + 
  geom_tile() +
  labs(title = "Proportion of Medals Won for Each Sport (Female)",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Proportion") +
  coord_fixed()
```


```{r}
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.12/bin/python3", required = TRUE)
```

```{python, include=FALSE}
import pandas as pd
import numpy as np
from palmerpenguins import load_penguins
from plotnine import *

# Read in data:
Athletes_df = pd.read_csv("/Users/tannerbessette/Desktop/SYE/athletes.csv")
Medals_df = pd.read_csv("/Users/tannerbessette/Desktop/SYE/medals.csv")


# Create Male and Female Datasets:
Male_Athletes_df = Athletes_df.query("gender == 'Male'")
Female_Athletes_df = Athletes_df.query("gender == 'Female'")


# Total Male and Female Medals for each sport:
# (learned reset_index from ChatGPT - helps to mimick summarise in R)
Male_Total_Medal_Counts = Medals_df.query("gender == 'M'").groupby("country_code").size().reset_index(name = "medal_totals")
Female_Total_Medal_Counts = Medals_df.query("gender == 'W'").groupby("country_code").size().reset_index(name = "medal_totals")


# MALE BAR PLOT:
# data manipulation (group_by, summarise):
Male_Athletes_Count = Male_Athletes_df.groupby("country_code").size().reset_index(name = "athlete_count")
# join data (looked ahead in textbook to ch. 22):
# (pd.merge is the join function, on is like by, how = join method(L/R))
Male_Athletes_Count_joined = pd.merge(Male_Athletes_Count, Male_Total_Medal_Counts, on="country_code", how="left")
# Only keep top 10 countries with most athletes:
Male_Athletes_Count_joined = Male_Athletes_Count_joined.query("athlete_count > 128")
# Create the Male bar plot with country's athletes and medals:
#Male_Athletes_Count.sort_values (essentially arrange -> do instead of reorder)
Male_Athletes_Count_joined = Male_Athletes_Count_joined.sort_values("athlete_count", ascending=False)
plot_1 = (
    ggplot() +
          geom_bar(aes(x = 'country_code', y = 'athlete_count'), fill = "lightblue", stat = "identity", data = Male_Athletes_Count_joined) +
          geom_bar(aes(x = 'country_code', y = 'medal_totals'), fill = "darkblue", stat = "identity", data = Male_Athletes_Count_joined) +
          labs(title = "Male Athletes and Medals by Country",
               x = "Country",
               y = "Male Athletes and Medal Counts") +
          theme_minimal()
)
plot_1.show()
```

```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# FEMALE BAR PLOT:
# data manipulation (group_by, summarise):
Female_Athletes_Count = Female_Athletes_df.groupby("country_code").size().reset_index(name = "athlete_count")
# join data (looked ahead in textbook to ch. 22):
# (pd.merge is the join function, on is like by, how = join method(L/R))
Female_Athletes_Count_joined = pd.merge(Female_Athletes_Count, Female_Total_Medal_Counts, on="country_code", how="left")
# Only keep top 10 countries with most athletes:
Female_Athletes_Count_joined = Female_Athletes_Count_joined.query("athlete_count > 177")
# Create the Male bar plot with country's athletes and medals:
# (arrange first):
Female_Athletes_Count_joined = Female_Athletes_Count_joined.sort_values("athlete_count", ascending=False)
plot_2 = (
    ggplot() +
          geom_bar(aes(x = 'country_code', y = 'athlete_count'), fill = "lightblue", stat = "identity", data = Female_Athletes_Count_joined) +
          geom_bar(aes(x = 'country_code', y = 'medal_totals'), fill = "darkblue", stat = "identity", data = Female_Athletes_Count_joined) +
          labs(title = "Female Athletes and Medals by Country",
               x = "Country",
               y = "Female Athletes and Medal Counts") +
          theme_minimal()
)
plot_2.show()
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Used ChatGPT for this, first step is sorting which sport will be the levels
# for the factor variable, second is assigning names for each level
Medals_df['Sport_Category'] = np.select(
    [
        Medals_df['discipline'].isin(["3x3 Basketball", "Basketball"]),
        Medals_df['discipline'].isin(["Badminton", "Table Tennis", "Tennis"]),
        Medals_df['discipline'].isin(["Boxing", "Taekwondo", "Wrestling", "Judo"]),
        Medals_df['discipline'].isin(["Swimming", "Marathon Swimming", "Diving", "Artistic Swimming"]),
        Medals_df['discipline'].isin(["Volleyball", "Beach Volleyball"]),
        Medals_df['discipline'].isin(["Cycling BMX Freestyle", "Cycling BMX Racing", 
                                      "Cycling Mountain Bike", "Cycling Road", "Cycling Track"]),
        Medals_df['discipline'].isin(["Water Polo", "Surfing", "Sailing"]),
        Medals_df['discipline'].isin(["Artistic Gymnastics", "Rhythmic Gymnastics", "Trampoline Gymnastics"]),
        Medals_df['discipline'] == "Athletics",
        Medals_df['discipline'].isin(["Canoe Slalom", "Canoe Sprint", "Rowing"]),
        Medals_df['discipline'] == "Equestrian",
        Medals_df['discipline'].isin(["Football", "Handball", "Hockey", "Rugby Sevens"]),
        Medals_df['discipline'].isin(["Archery", "Shooting", "Fencing"]),
        Medals_df['discipline'].isin(["Triathlon", "Modern Pentathlon"]),
        Medals_df['discipline'] == "Weightlifting",
        Medals_df['discipline'] == "Skateboarding",
        Medals_df['discipline'] == "Breaking",
        Medals_df['discipline'] == "Sport Climbing",
        Medals_df['discipline'] == "Golf"
    ],
    [
        "Basketball",
        "Racket Sports",
        "Combat Sports",
        "Swimming",
        "Volleyball",
        "Cycling",
        "Water Sports",
        "Gymnastics",
        "Athletics",
        "Canoe/Rowing",
        "Equestrian",
        "Team Sports",
        "Precision Sports",
        "Athlons",
        "Weightlifting",
        "Skateboarding",
        "Breakdancing",
        "Climbing",
        "Golf"
    ],
    default="Other"  # default if any sport doesn't match
)

# Convert 'Sport_Category' to a categorical type
Medals_df['Sport_Category'] = Medals_df['Sport_Category'].astype('category')
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Medal_Counts = Medals_df.groupby(['country_code', 'Sport_Category']).size().reset_index(name='count').sort_values(by='count', ascending=False)
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# Create a total medal count variable, and only keep countries 
# with medals in at least 1 sport
total_counts = Medal_Counts.groupby('country_code').size().reset_index(name='total_count')
Medal_Counts = pd.merge(Medal_Counts, total_counts, on = "country_code",
how = "left")
Medal_Counts = Medal_Counts.query("total_count > 1")
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
overall_countries_to_keep = ['AUS', 'BRA', 'CAN', 'CHN', 'ESP', 'FRA', 'GBR', 'GER', 'HUN', 'ITA', 'JPN', 'KOR', 'NED', 'NZL', 'USA']

Overall_Medal_Counts = Medal_Counts.query("country_code in @overall_countries_to_keep").query("count > 0")

plot_3 = (
    ggplot(Overall_Medal_Counts, aes(x='Sport_Category', 
                             y='country_code', 
                             fill='count')) +  
    geom_tile() +
    labs(title="Medal Count by Country and Sport",
         x="Sport",
         y="Country") +
    theme_minimal() +
    theme(axis_text_x=element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_fixed() +
    scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count")
)
plot_3.show()
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Male_Medal_Counts = Medals_df.query("gender == 'M'").groupby(['country_code', 'Sport_Category']).size().reset_index(name = 'count').sort_values(by = 'count', ascending = False)
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
total_male_counts = Male_Medal_Counts.groupby('country_code').size().reset_index(name='total_count')
Male_Medal_Counts = pd.merge(Male_Medal_Counts, total_male_counts, on = "country_code", how = "left")
Male_Medal_Counts = Male_Medal_Counts.query("total_count > 1")
```

```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
male_countries_to_keep = ['AUS', 'CHN', 'ESP', 'FRA', 'GBR', 'GER', 'HUN', 'IRI', 'ITA', 'JPN', 'KOR', 'NED', 'USA', 'UZB']

Updated_Medal_Counts = Male_Medal_Counts.query("country_code in @male_countries_to_keep").query("count > 0")

plot_4 = (
    ggplot(Updated_Medal_Counts, aes(x='Sport_Category', 
                             y='country_code', 
                             fill='count')) +  
    geom_tile() +
    labs(title="Medal Count by Country and Sport (Male)",
         x="Sport",
         y="Country") +
    theme_minimal() +
    theme(axis_text_x=element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_fixed() +
    scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count")
)
plot_4.show()
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Female_Medal_Counts = Medals_df.query("gender == 'W'").groupby(['country_code', 'Sport_Category']).size().reset_index(name = 'count').sort_values(by = 'count', ascending = False)
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
total_female_counts = Female_Medal_Counts.groupby('country_code').size().reset_index(name='total_count')
Female_Medal_Counts = pd.merge(Female_Medal_Counts, total_female_counts, on = "country_code", how = "left")
Female_Medal_Counts = Female_Medal_Counts.query("total_count > 1")
```

```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
female_countries_to_keep = ['AUS', 'BRA', 'CAN', 'CHN', 'FRA', 'GBR', 'GER',  'ITA', 'JPN', 'KOR', 'NED', 'NZL', 'USA']

Updated_Female_Medal_Counts = Female_Medal_Counts.query("country_code in @female_countries_to_keep").query("count > 0")

plot_5 = (
    ggplot(Updated_Female_Medal_Counts, aes(x='Sport_Category', 
                             y='country_code', 
                             fill='count')) +  
    geom_tile() +
    labs(title="Medal Count by Country and Sport (Female)",
         x="Sport",
         y="Country") +
    theme_minimal() +
    theme(axis_text_x=element_text(angle=90, vjust=0.5, hjust=1)) +
    coord_fixed() +
    scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count")
)
plot_5.show()
```

```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Male_prop = Male_Medal_Counts
Male_prop['total_sport_medals'] = Male_prop.groupby('Sport_Category')['count'].transform('sum')
Male_prop['prop_sport_medals'] = Male_prop['count'] / Male_prop['total_sport_medals']
```

```{python, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Female_prop = Female_Medal_Counts
Female_prop['total_sport_medals'] = Female_prop.groupby('Sport_Category')['count'].transform('sum')
Female_prop['prop_sport_medals'] = Female_prop['count'] / Female_prop['total_sport_medals']
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Updated_Male_Prop = Male_prop.query("country_code in @male_countries_to_keep").query("count > 0")

plot_6 = (
    ggplot(Updated_Male_Prop, aes(x='Sport_Category', 
                                  y='country_code', 
                                  fill='prop_sport_medals')) +  
    geom_tile() +
    labs(title="Proportion of Medals Won for Each Sport (Male)",
         x="Sport",
         y="Country") +
    theme_minimal() +
    theme(axis_text_x=element_text(angle=90, vjust=0.5, hjust=1)) +
    scale_fill_gradient(low="lightblue", high="darkblue", name="Medal Proportion") +
    coord_fixed()
)
plot_6.show()
```


```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
Updated_Female_Prop = Female_prop.query("country_code in @female_countries_to_keep").query("count > 0")

plot_7 = (
    ggplot(Updated_Female_Prop, aes(x='Sport_Category', 
                                  y='country_code', 
                                  fill='prop_sport_medals')) +  
    geom_tile() +
    labs(title="Proportion of Medals Won for Each Sport (Female)",
         x="Sport",
         y="Country") +
    theme_minimal() +
    scale_fill_gradient(low="lightblue", high="darkblue", name="Medal Proportion") +
    theme(axis_text_x=element_text(angle=90, vjust=1, hjust=1)) +
    coord_fixed()
)
plot_7.show()
```





## SYE Goals

- Learn Python for data science
- Go through a Python data science textbook and explore Positron IDE
- Compare R and Python functions side by side
- Be able to identify strengths of both languages, and combine them in a project

## What is R? What is Python?

- Both R and Python are coding languages
- R is specifically designed for statistics/data analysis
- Python is utilized for a wider variety of things, including software 
development, web applications, machine learning, and data science

## Olympics Project Overview

- Split the semester into two main projects: 2024 Paris Olympics and NBA
- First: Wrangle data and generate plots in R
- Next: replicate everything in Python, following along with a textbook called 
"Python for Data Science"

## Some Olympics Project Plots

**Female Athletes Barplot:**

- Light blue represents number of athletes each country sent to the Olympics

- Dark blue represents the number of medals won

```{r, warning=FALSE, message=FALSE, fig.width=3, fig.height=2, fig.align="center"}
Female_Athletes_Count <- Female_Athletes_df %>%
  group_by(country_code) %>%
  summarise(athlete_count = n()) %>%
  arrange(desc(athlete_count))

# join the medals counts with the athletes count
Female_Athletes_Count <- left_join(
  Female_Athletes_Count, Female_Total_Medal_Counts, by = "country_code")

# Calculate max country counts and visualize the top 10 countries:
Female_Athletes_Count <- Female_Athletes_Count %>%
  filter(athlete_count > 177)

# Create the plot:
Female_Athletes_Count_barplot <- ggplot(data = Female_Athletes_Count) +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = athlete_count), fill = "lightblue", stat = "identity") +
          geom_bar(aes(x = reorder(country_code, -athlete_count), y = medal_totals), fill = "darkblue", stat = "identity") +
          labs(title = "Female Athletes and Medals by Country",
               x = "Country",
               y = "Female Athletes and Medal Counts") +
          theme_minimal() +
          scale_fill_viridis_c() +
          theme(plot.title = element_text(size = 7, hjust = 0.5),
                axis.title.x = element_text(size = 6),
                axis.title.y = element_text(size = 6),
                axis.text.x = element_text(size = 5),
                axis.text.y = element_text(size = 5),
                legend.position = "none")

print(Female_Athletes_Count_barplot)
```

## Some Olympics Project Plots

**Male Medals Heatmap:**

- Each heatmap square corresponds to one country and one sport  

- The darkness of the color represents the number of medals that country's male athletes won in that sport

```{r, warning=FALSE, message=FALSE, fig.width=3, fig.height=2, fig.align="center"}
# Only keep countries who got at least 10 total medals:
# (because heatmap was getting sloppy)
# group by our new Sport_Category variable before counting number of medals
Male_Medal_Counts <- Medal_Counts  %>%
  filter(gender == "M") %>%
  group_by(country_code, Sport_Category) %>%
  mutate(count = n()) %>%
  arrange(desc(count))

Male_Medal_Counts <- Male_Medal_Counts %>%
  group_by(country_code) %>%
  mutate(total_count = n()) %>%
  filter(total_count >= 10)

# Heatmap 
ggplot(Male_Medal_Counts, aes(x = Sport_Category, 
                              y = reorder(country_code, total_count), 
                              fill= count)) + 
  geom_tile() +
  labs(title = "Medal Count by Country and Sport (Male)",
               x = "Sport",
               y = "Country") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  scale_fill_gradient(low="lightblue", high="darkblue", name = "Medal Count") +
  coord_fixed() +
  theme(plot.title = element_text(size = 7, hjust = 0.5),
                axis.title.x = element_text(size = 6),
                axis.title.y = element_text(size = 6),
                axis.text.x = element_text(size = 5),
                axis.text.y = element_text(size = 5),
                legend.key.size = unit(0.3, "cm"),
                legend.text = element_text(size = 4), 
                legend.title = element_text(size = 4))
```

## R vs. Python Code Comparison

- Here is how I performed the joining of the male athletes and male medals datasets before creating the bar plot using R:

```{r, echo = TRUE}
Male_Athletes_Count <- 
  left_join(Male_Athletes_Count,
            Male_Total_Medal_Counts, 
            by = "country_code")
```

- And here is how I did that in Python:

```{python, echo = TRUE}
Male_Athletes_Count_joined = pd.merge(
  Male_Athletes_Count, Male_Total_Medal_Counts,
  on="country_code", how="left")
```

## R vs. Python Code Comparison

- Before making the male bar plot, I had to group by country and add up all the athletes for each country.
- Here is how I did it in R:

```{r, echo = TRUE, results = 'hide'}
Male_Athletes_Count <- Male_Athletes_df %>%
  group_by(country_code) %>%
  summarise(athlete_count = n())
```

- And here is how I did it in Python:
```{python, echo = TRUE, output = FALSE}
Male_Athletes_Count = Male_Athletes_df.groupby(
"country_code").size().reset_index(name="athlete_count")
```

## NBA Project

- The second project I worked on was predicting whether NBA players made the all-star game
based on their stats
- The dataset included all player statistics each season from 1947 to present
- First part of the project: wrangle and plot in R
- Second part: perform a statistical learning method (KNN) in Python

## NBA Project Results

- Interestingly, when making the exploratory plots, one of the variables that looked the most correlated with players making the all-star game was having a high average turnovers
- The variables I used in the KNN model were: points, rebounds, assists, steals, field goals attempted, and turnovers
- Final KNN model accuracy was 88.22%

## Turnovers and Three Pointers as Predictors of All-Star

```{r, warning=FALSE, message=FALSE, include=FALSE, echo=FALSE, results='hide'}
library(tidyverse)
library(dplyr)
player_averages <- 
  read_csv("/Users/tannerbessette/Desktop/Python_Textbook/NBA_Project/Player_Per_Game.csv")

all_stars <- read_csv("/Users/tannerbessette/Desktop/Python_Textbook/NBA_Project/All_Star_Selections.csv")

player_totals <- 
  read_csv("/Users/tannerbessette/Desktop/Python_Textbook/NBA_Project/Player_Totals.csv")
```

```{r, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# create an all-star variable with 1 = made all-star and 0 = didn't make it:
all_stars <- all_stars |>
  mutate(all_star = 1)

# Join player averages data with the all_stars data:
joined_all_star_data <- left_join(player_averages, all_stars, 
                                  by = c("season", "player"))

# remove variables that we definitely won't need:
joined_all_star_data <- joined_all_star_data |>
  dplyr::select(-c("birth_year", "lg.x", "lg.y", "team"))

# fix the all-star variable to be 0s and 1s not 1s and NAs
joined_all_star_data <- joined_all_star_data |>
  mutate(all_star = case_when(
    is.na(all_star) ~ 0,
    all_star == 1 ~ 1))
```

```{r, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# specify the seasons:
joined_all_star_data <- joined_all_star_data |>
  filter(season > 1999) |>
  filter(season != 2025)

# specify the number of starts each player must have (at least half of the season):
joined_all_star_data <- joined_all_star_data |>
  filter(gs > 41)
```

```{r, warning=FALSE, message=FALSE, fig.width = 5, fig.height = 3}
library(gridExtra)
library(viridis)

# Three pointers made per game:
plot1 <- ggplot(joined_all_star_data, aes(x = as.factor(all_star), 
                                 y = x3p_per_game)) +
    geom_boxplot() +
    labs(x = "All Star? (0 = No, 1 = Yes)",
         y = "3 Pointers Made per Game",
         title = "3 Pointers as a Predictor of All-Star") +
    theme_minimal() +
    scale_fill_viridis_d(name = "All Star", option = "D") +
    theme(plot.title = element_text(size = 8, hjust = 0.5, face = "bold"),
                axis.title.x = element_text(size = 7),
                axis.title.y = element_text(size = 7),
                axis.text.x = element_text(size = 6),
                axis.text.y = element_text(size = 6))

# Turnovers per game:
plot2 <- ggplot(joined_all_star_data, aes(x = as.factor(all_star), 
                                 y = tov_per_game)) +
    geom_boxplot() +
    labs(x = "All Star? (0 = No, 1 = Yes)",
         y = "Turnovers per Game",
         title = "Turnovers a Predictor of All-Star") +
    theme_minimal() +
    scale_fill_viridis_d(name = "All Star", option = "D") +
    theme(plot.title = element_text(size = 8, hjust = 0.5, face = "bold"),
                axis.title.x = element_text(size = 7),
                axis.title.y = element_text(size = 7),
                axis.text.x = element_text(size = 6),
                axis.text.y = element_text(size = 6))

# Combine the plots side by side
grid.arrange(plot1, plot2, ncol=2)
```

## Confusion Matrix
```{python, include=FALSE, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
import math
from sklearn.neighbors import KNeighborsClassifier
import sklearn.metrics

# read in the data we wrangled previously
joined_all_star_data = pd.read_csv("/Users/tannerbessette/Desktop/Python_Textbook/NBA_Project/joined_all_star_data.csv")

# set the seed
np.random.seed(42)

# .sample in Python appears to be the same as slice_sample in R
train_sample = joined_all_star_data.sample(n = 2852)

# below is what we are using instead of an anti-join
# basically does a left join but creates a merge variable that
# indicates whether a row was in both original datasets, the left,
# or the right. Only keep in left, and remove the variable afterwards

columns_to_join_on = joined_all_star_data.columns.tolist()
test_sample = joined_all_star_data.merge(train_sample, on=columns_to_join_on, how='left', indicator=True)

test_sample = test_sample[test_sample['_merge'] == 'left_only'].drop(columns=['_merge'])

# to specify the columns to use for knn model, create a list, then
# select just those columns from train and test datasets
columns_to_keep = ['pts_per_game', 'trb_per_game', 'ast_per_game', 
                     'stl_per_game', 'fga_per_game', 'tov_per_game']

train_small = train_sample[columns_to_keep]
test_small = test_sample[columns_to_keep]

# Create vectors:
train_cat = train_sample['all_star']
test_cat = test_sample['all_star']

# Find a good number of nearest neighbors (sqrt of train size):
math.sqrt(2852) # use 53 nearest neighbors

# Specify 53 neighbors:
knn = KNeighborsClassifier(n_neighbors = 53)

# Fit the knn model
knn.fit(train_small, train_cat)

# Make predictions for test data
knn_mod = knn.predict(test_small)
```

- **Confusion matrix with the results of my KNN model**
- Top left number, 559, gives us the number of non-all-stars that our knn model correctly predicted as non-all-stars
- Bottom right number $\rightarrow$ number of all-stars that our model correctly predicted as all-stars. 
- Bottom left $\rightarrow$ non-all-stars that our model predicted to actually be all-stars
- Top right $\rightarrow$ players that our model predicted to not be all-stars but were actually all-stars

```{python, echo = TRUE, class.source="small-code", class.output="small-output"}
# Create and display the confusion matrix
conf_matrix = sklearn.metrics.confusion_matrix(
  test_cat, knn_mod)
print(conf_matrix)
```

## Thoughts on Learning Python for Data Science

- I am surprised that Python is used so much for data science outside of SLU
- R seems more intuitive and organized, especially when it comes to data wrangling
- As an introduction to coding or a machine learning class, I believe
Python makes more sense to learn, but I'm glad our department teaches R for
DATA234

## Conclusion

- In my opinion, R is easier to learn for data science, especially for beginner coders
- Both languages have similar math/statistical capabilities
- For data wranging and plotting, R is great!
- Python may be better (or at least equal) in terms of stat. learning


